Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-16 paper/main.tex
restructure methods part
- make paragraphs into subsubsections
- shift figure inside of section

\paragraph{Direct Comparison}
\label{paragraph:direct-comparison}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the positions
calculated from the computational results.
Each cell contains a collection of vertices.
Given $M$ cells with $N$ vertices each and dimension $d$, the total information can be stored in an
array $\nu_{i,j,k}$ of shape $(M,N,2)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays within a suitable cost function~\cite{Wang2020}.
In our approach we utilize a least squares approach given by

@@

\paragraph{Comparison across Generations}
The preceding algorithm is performant and effective but can not be applied in the event of a
division process.
Due to possibly differing division times, a cell which is present in the simulation might not yet
have a corresponding cell in the experimental data at this time point yet (or vice-versa) (see
Figure~\ref{fig:mask-difference-metric} (E)).
In order to still be able to compare these datapoints, we require additional methods which can be
used for comparisons across division events.
\textbf{[Es bleibt unklar, was du hier verhandeln möchtest.]}
\textit{Jetzt klarer?}

@@

@@

\paragraph{Testing the Comparison Algorithm with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ with $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area in time,
which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (F) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.
\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar. Es könnte nützlich sein, vor der Application Sektion, eine Zusammenfassung zu setzen.]}

@@

\begin{itemize}
    \item go from simple to more complex: single-cell, many cells, with division
    \item explain data in a generalized way; detailed explanations within subsections
\end{itemize}

>>>>

\subsubsection{Direct Comparison}
\label{paragraph:direct-comparison}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the positions
calculated from the computational results.
Each cell contains a collection of vertices.
Given $M$ cells with $N$ vertices each and dimension $d$, the total information can be stored in an
array $\nu_{i,j,k}$ of shape $(M,N,2)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays within a suitable cost function~\cite{Wang2020}.
In our approach we utilize a least squares approach given by

@@

\subsubsection{Imaging-based Comparison Algorithm}
The preceding algorithm is performant and effective but can not be applied in the event of a
division process.
Due to possibly differing division times, a cell which is present in the simulation might not yet
have a corresponding cell in the experimental data at this time point yet (or vice-versa) (see
Figure~\ref{fig:mask-difference-metric} (E)).
In order to still be able to compare these datapoints, we require additional methods which can be
used for comparisons across division events.
\textbf{[Es bleibt unklar, was du hier verhandeln möchtest.]}
\textit{Jetzt klarer?}

@@

\paragraph{Testing the Comparison Algorithms with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ with $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area in time,
which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (F) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.
\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar. Es könnte nützlich sein, vor der Application Sektion, eine Zusammenfassung zu setzen.]}

@@

@@
